Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-24 bbbq_article.tex
Feedback Frans, thanks @fransbianchi

Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology was determined 
using PureseqTM.
From the topology and the known location of the SNP, 
we score the location (i.e. TMH or soluble protein region) 
where the change occurred.

@@

[RJCB: I suggest to move this section to the Supplementary Materials]

@@

The R package 'tmhmm' was developed to do the similar topology
predictions as our earlier study (that used 'TMHMM'), yet in an automated way.
'TMHMM' has a restrictive software license (\cite{krogh2001predicting}) and allows a user
to download a pre-compiled executable after confirmation he/she
is in academia. The R package respects this restriction
and allows the user to install and use TMHMM from within R,
as done in this study.
'tmhmm' has been submitted to and is accepted by CRAN.

@@

To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
we reanalyzed published peptide elution studies.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively),
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
From these findings, it is robustly predicted that
epitopes derived from TMHs are presented by B cells [For Frans: please check
that the article by Bergseng, 2015, 'Different binding motifs of the celiac 
disease-associated HLA molecules DQ2. 5, DQ2. 2, and DQ7. 5 revealed by 
relative quantitative proteomics of endogenous peptide repertoires'
is appropriate. As far as I can see, it is, but I may overlook something. Richel].

@@

We recently showed that peptides derived from TMHs of human proteins 
were predicted to be over-presented in MHC-I. 
Here, we wondered if this finding would be general 
and if peptides stemming from TMHs of microbial pathogens 
would also be over-presented.
Figure \ref{fig:1} shows the percentages of MHC-I epitopes 
predicted to be overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc1}) 
for the exact TMH and epitope counts.
We found that also for pathogen epitopes derived from TMHs are
over-presented. 
The extent of the over-presentation is similar between the haplotypes:
if a haplotype if prone to present TMH-derived epitopes from
human proteins, it is also prone to present such epitopes of
bacterial and viral proteins.

@@

Where TMH-derived epitopes are over-presented by MHC-I,
we wondered if this finding would also generalize to MHC-II.
Figure \ref{fig:2} shows the percentages of MHC-II epitopes overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc2}) 
for the TMH and epitope counts.
We show that MHC-II has the same pattern of over-presentation
as MHC-I, yet this over-presentation is more extreme.
Also for MHC-II, we observe that the extent the which TMH-derived
epitopes are presented is similar between the haplotypes, for
a human, bacterial and viral proteome.

@@

In the final set of experiments, 
we addressed the question whether TMHs 
would be more evolutionary conserved than soluble proteins regions 
by comparing the occurrences of SNPs in the genome coding 
for TMHs and soluble protein regions within membrane proteins.
We obtained 1,129 [For Frans: I would enjoy to write 'approx 1k' here, would that
be acceptable? I do refer to a table with the exact values
later on anyways. Richel] gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs (hence 
2,237 proteins are predicted to be MAPs).
We obtained 37,630 SNPs that resulted in an
amino acids substitution, of which 21,024 were located 
in predicted TMPs.
See tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for an overview of all amounts.

Per protein, we calculated two percentages: 
(1) the percentage of the protein coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without MAPs,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs 
and soluble protein domains.
This analysis gives the first hint that SNPs 
occur more often in soluble domains than in TMHs.

To statistically test whether SNPs occur 
just as likely in soluble domains as
in TMHs, we first calculated the number of SNPs that are expected to be
found in TMHs of TMPs by chance, $E(n_{SNP\_TMH})$,
equals the number of SNPs in TMHs, $n_{SNP} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH} = 18.66\%$,
$E(n_{SNP\_TMH}) = n_{SNP} \cdot f_{TMH} \approx 4056$.

@@

The actually number of SNPs found in TMHs
was $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4056$.
To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails,
$n$ equals the number of SNPs, which is 21,738. The probability of success
for TMP $i$, $p\_i$, is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals [RJCB: recalculate]
$1.6201 \cdot 10^{-9}$.

@@

We conclude that,
when randomly selecting SNPs, 
for 100 SNPs found in soluble regions of multi-spanners,
one finds 89.6 SNPs in TMHs, as shown by 
figure \ref{fig:conservation_per_spanner}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented in vivo, we conclude that these are additionally
presented more often than expected chance 
alone (figures \ref{fig:1} and \ref{fig:2}).
This in silico finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans and two pathogens.

@@

We conclude that mutations are more likely to occur
in soluble regions of proteins, than in TMHs.
When selecting SNPs at random, per 1000 SNPs found in a soluble domain, 
one finds on average 918 SNPs in TMHs.
This effect is present in TMPs with two or more TMHs, 
but not in single-spanners.

@@

Our understanding of the immune system and epitope prediction is 
important to fight disease. By this paper, we suggest that epitopes
derived from TMHs are needlessly overlooked and may inspire to investigate
new potential targets for combating pathogens.

\subsection{Elution studies}

@@

From elution studies and in silico predictions, 
we suggest that TMH-derived epitopes are presented in vivo.
The biochemical intracellular pathway to achieve this,
however, is still unknown.

@@

An additional unknown is, although our findings are robust,
whether this is a false positive,
due to additive methodological artifacts.
An obvious artifact is that the elution studies used 
will contain some unavoidable sequencing errors.
Additionally, we assume that the epitopes naturally presented
are derived from a cleanly translated proteome.
Two known biological phenomena that violate this assumption
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion (\cite{delong2016pathogenic}),
two processes that both result in peptide fragments
that did not originate from genome directly.
Also, the TMH prediction software is imperfect.
This imperfection, however, appears to be of limited importance,
as two different topology prediction software have similar results.
Although those a couple of percent of epitopes that 
can be mapped directly to the proteome,
there is methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy. This means that the number of TMH-derived epitopes
is underestimated.
in vivo work in the lab will be needed to resolve this finding.

@@

% \paragraph{MHC-II is expected to present TMHs}

@@

In general, on would hope that evolutionary selection results in
an immune system that as most attentive for loci that are
essential for a virus, as these will be most conserved.
In SARS-CoV-2, for example, there is preliminary evidence that the strongest
selection pressure is upon residues that changes its 
virulence (\cite{velazquez2020positive}).
These loci, however, only account for a small part of a pathogen's proteome.
Additionally, these essential parts differ widely between pathogens.
Because of this scarcity and variance in targets, 
one can imagine that the human immune system 
is not tailored to detect these sites, 
as hinted by upon by a extensive study on influenza (\cite{han2019individual}),
where it was found that the selection pressure
exerted by the immune system was only limited.

@@

TMHs, on the other hand, also have their function constraints, 
yet can occur multiple time a pathogen's proteome.
One can safely assume a pathogen's proteome contains multiple TMHs.
Therefore, it may be beneficial for the host
if its immune system would be more attentive towards TMHs.
And maybe this has already happened: MHC-I already detects hydrophobic
peptides. This feature, however, may also be caused by selection
to detect hydrophobic regions in the soluble proteins of pathogens.
It is unknown, when focusing on TMHs only, if a signal of selection
can be detected.
We speculate that the human immune system is more attentive 
towards TMHs, as it is a general structure, that is evolutionarily conserved.

We aimed our evolutionary experiment at TMHs, because these can
be predicted well from a protein structure,
are common structures and are present in all pathogens. 
We could have done the same experiment on beta-turn,
as also these can be predicted well (\cite{petersen2010netturnp}),
are common structures and are present in all pathogens.
We speculate that the human immune system may also be more attentive 
towards beta turns for this reason.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster. 
Additionally, we would like to thank Sci-Hub (\cite{himmelstein2018sci})
for allowing us to read paywalled articles while working from home.
[GVDB: 
  Should we include Maxim? 
  He did the alignment of Mycobacterium tuberculosis. 
  Let’s also ask Frans.
]

@@

To retrieve the data from the NCBI databases the
\verb;rentrez; R package (\cite{rentrez}) was used
that calls the NCBI website's API. To provide for a 
stable user experience for all users, 
this API limits the user to 3 calls per second.
Additionally, the API splits the result of a bigger
query into multiple pages, each of which needs one API call.
We wrote the \verb;sprentrez; package (\cite{sprentrez}) to provide for 
bigger queries of multiple (and delayed) API calls.}

>>>>

Per SNP, the 'protein' NCBI database was queried for the
protein sequence.
Of each protein sequence, the protein topology was determined 
using PureseqTM.
Using the protein topology SNPs were scored to be in or outside trans membrane domains.

@@

[RJCB: I suggest to move this section to the Supplementary Materials]
\frans{I agree move to supp}

@@

The R package 'tmhmm' was developed to do the similar topology
predictions as our earlier study (that used 'TMHMM'), yet in an automated way.
'TMHMM' has a restrictive software license (\cite{krogh2001predicting}) and allows a user
to download a pre-compiled executable after confirmation he/she
is in academia. The R package respects this restriction
and allows the user to install and use TMHMM from within R,
as done in this study.
'tmhmm' has been submitted to and is accepted by CRAN. \frans{wat is CRAN}

@@

To confirm that peptides stemming from TMHs were presented in MHC-I and MHC-II,
peptide elution studies were reanalyzed.
Table \ref{tab:elution} shows the percentage of epitopes derived 
from a TMH
found in MHC-I and MHC-II elution 
studies (\cite{schellens2015comprehensive} and \cite{bergseng2015different} respectively),
for the two topology prediction tools TMHMM and 
PureseqTM (\cite{krogh2001predicting} and \cite{wang2019efficient} respectively). 
Regardless of the prediction, 
at least 100 epitopes were predicted to be derived from a TMH. 
From these findings, it is robustly predicted that
epitopes derived from TMHs can and are presented on both MHC-I and MHC-II.

@@

We recently showed that peptides derived from TMHs of human proteins 
were predicted to be over-presented on HLA-A and HLA-B. 
Here, we wondered if this finding would be general 
and if peptides stemming from predicted TMHs of microbial pathogens 
would also be over-presented on HLA-A and HLA-B.We therefor analysed all superhaplotypes of the HLA-A and HLA-B class see Figure \ref{fig:1} and found that similar as for the human proteome the percentages of MHC-I epitopes predicted to be overlapping with TMHs for our human, viral and bacterial proteome was higher as expected by TMH abundance.
See the supplementary materials (table \ref{tab:tmh_binders_mhc1}) 
The extent of the over-presentation between the various different HLA supertypes is similar for the probed proteomes, which confirms that the hydrophopbicity of the MHC binding cleft is determinant.

@@

We next wondered if this over-representation would also generalize to MHC-II. \frans{ you say MHC-II but this is very general I think you have to explain the choice for these haplotypes.}
Unlike MHC-I where only one variable chain determines the landscape of the epitope binding pocket allowing for the classification of 5 HLA-A and 8 HLA-B superhaplotypes with .... variation in peptide binding repertoire, the MHC-II peptide binding groove is determined by both the alpha and beta chain. For HLA DQ both the alpha and beta chain is highly polymorphic  ..... \frans{why is HLA-DP not predicted?}
We took the alleles that represent percentage of the occurrence int he overall human population according to ... et al. 
Figure \ref{fig:2} shows the percentages of MHC-II epitopes overlapping 
with TMHs for our human, viral and bacterial proteome.
See the supplementary materials (table \ref{tab:tmh_binders_mhc2}) 
for the TMH and epitope counts.
We show that MHC-II has the same pattern of over-presentation
as MHC-I, yet this over-presentation is more pronounced. \frans{we should explain this does it like shown for mhc-I also linearly correlates with the hydrophopbicity of the MHC-II binding celft?}
Also for MHC-II, we observe that the extent to which TMH-derived
epitopes are presented is similar between the haplotypes, for
a human, bacterial and viral proteome.

@@

In general the occurrence and expression levels of membrane proteins are low, therefor we addressed the question whether there is an evolutionary advantage for presenting TMHs. To this end we looked for conservation the conservation in TMH vs soluble regions by comparing the occurrences of SNPs in the genome coding for TMHs versus soluble protein regions within membrane proteins.
We obtained >1000 gene names associated with the phrase 'membrane protein',
which are genes for both membrane-associated proteins (MAPs, no TMH) and 
transmembrane proteins (TMPs, at least one TMH).
These genes are linked to 4,780 protein isoforms, 
of which 2,553 are predicted to be TMPs (hence 
2,237 proteins are predicted to be MAPs).
We obtained 37,630 SNPs that resulted in an
amino acids substitution, of which 21,024 were located 
in predicted TMPs.
See tables \ref{tab:ncbi_counts_1} and \ref{tab:ncbi_counts_2} 
for an overview of all amounts.

Per protein, we calculated two percentages: 
(1) the percentage of the protein coding for TMHs, 
and (2) the percentage of SNPs occurring in TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analyses on the data twice,
one with and one without MAPs,
and added a 95\% confidence interval for both of these regressions.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would be expected if SNPs occur just as likely within TMHs 
and soluble protein domains.
This analysis gives the first hint that SNPs 
occur more often in soluble domains than in TMHs.\frans{waar vind ik dan nu het verschil tussen met en zonder MAPs je zou hopen dat er geen verschil is toch omdat het aandeel TMH in de analyse met MAPS ook naar beneden gaat}
To statistically test whether SNPs occur 
just as likely in soluble domains as
in TMHs, we first calculated the number of SNPs that are expected to be
found in TMHs of TMPs by chance, $E(n_{SNP\_TMH})$,
equals the number of SNPs in TMHs, $n_{SNP} = 21,738$, 
multiplied by the fraction of the TMPs that are TMH, $f_{TMH} = 18.66\%$,
$E(n_{SNP\_TMH}) = n_{SNP} \cdot f_{TMH} \approx 4056$.

@@

The actual number of SNPs found in TMHs
was $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4056$.
To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails,
$n$ equals the number of SNPs, which is 21,738. The probability of success
for TMP $i$, $p\_i$, is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals [RJCB: recalculate]
$1.6201 \cdot 10^{-9}$.

@@

We conclude that,
when randomly selecting SNPs, 
for 100 SNPs found in soluble regions of multi-spanners,
one finds 89.6 SNPs in TMHs, as shown by \frans{it says 91,1 in fig 4}
figure \ref{fig:conservation_per_spanner}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}\frans{why comes this section before discussion?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented in vivo, we conclude that these are additionally
presented more often than expected by chance 
alone (figures \ref{fig:1} and \ref{fig:2}).
This in silico finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans, bacteria and viruses.

@@

We conclude that mutations are more likely to occur
in soluble regions of proteins, than in TMHs.
When selecting SNPs at random, per 1000 SNPs found in a soluble domain, 
one finds on average 918 SNPs in TMHs.
This effect is more pronounced in multi spanners likely as these TMHs more often serve additional functions, besides spanning the lipid bilayer.

@@

Our understanding of the immune system and epitope prediction is 
important to fight disease. By this paper, we suggest that epitopes
derived from TMHs are needlessly overlooked and may inspire to investigate
new potential targets for combating pathogens.
\frans{ik zie hier verder geen argumentatie voor deze statement}
\subsection{Elution studies}

@@

From elution studies and in silico predictions, 
we suggest that TMH-derived epitopes are presented and cross-presented in vivo. How integral membrane proteins are presented and cross-presented on MHC complexes is however unclear. Most prominently because the fundamental principles of trans locating integral membrane proteins from there hydrophobic environment into the vacuole lumen and subsequent processing prior to loading on MHC complexes is unresolved.

@@

An additional unknown is, although our findings are robust,
whether this is a false positive,
due to additive methodological artifacts.
An obvious artifact is that the elution studies used 
will contain some unavoidable sequencing errors.
Additionally, we assume that the epitopes naturally presented
are derived from a cleanly translated proteome.
Two known biological phenomena that violate this assumption
are defective ribosomal products (DRiPs) (\cite{yewdell1996defective}) 
and peptide fusion (\cite{delong2016pathogenic}),
two processes that both result in peptide fragments
that did not originate from genome directly.
Also, the TMH prediction software is imperfect.
This imperfection, however, appears to be of limited importance,
as two different topology prediction software algorithms gave similar results.
Although those a couple of percent of epitopes that 
can be mapped directly to the proteome,
there is methodological bias caused by mass spectronomy,
as hydrophobic fragments are detected less 
by mass spectronomy. This means that the number of TMH-derived epitopes
is underestimated.
in vivo work in the lab will be needed to resolve this finding.

@@

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevnet zijn de haplotypen die veel TMH presenteren?}

The over representation of TMH on MHC-II haplotypes such as HLADQA!*0101/DQB*0501 is more pronounced as for MHC-I with a third of the predicted binders from the human and MtB proteome stem from TMH  human reference proteome and MtB reference proteome to be presented on MHC-II. \frans{ik zoek nog naar associaties}

@@

In general, on would hope that evolutionary selection results in
an immune system that as most attentive for loci that are
essential for a virus, as these will be most conserved.
In SARS-CoV-2, for example, there is preliminary evidence that the strongest
selection pressure is upon residues that changes its 
virulence (\cite{velazquez2020positive}).
These loci, however, only account for a small part of a pathogen's proteome.
Additionally, these essential parts differ widely between pathogens.
Because of this scarcity and variance in targets, 
one can imagine that the human immune system 
is not tailored to detect these sites, 
as hinted upon by an extensive study on influenza (\cite{han2019individual}),
where it was found that the selection pressure
exerted by the immune system was a non pro founding factor.

@@

TMHs, on the other hand, also have their functional constraints, 
and make up ... of a human, bacterial or viral proteome.
One can safely assume a proteome contains many TMHs,
therefore it may be beneficial for the host
if its immune system would be more attentive towards TMHs.
And maybe this has already happened: MHC-I already detects hydrophobic
peptides. This feature, however, may also be caused by selection
to detect hydrophobic regions in the soluble proteins of pathogens.
It is unknown, when focusing on TMHs only, if a signal of selection
can be detected.\frans{ik snap deze redenatie niet?}
We speculate that the human immune system is more attentive 
towards TMHs, as it is a general structure, that is evolutionarily conserved.

We aimed our evolutionary experiment at TMHs, because these can
be predicted well from a protein structure,
are common structures and are present in all pathogens. \frans{ dit doen we toch omdat we vinden dat er een preferred binding affinity is voor hydrophobe peptides?}
We could have done the same experiment on beta-sheets,
as also these can be predicted well (\cite{petersen2010netturnp}),
are common structures and are present in all pathogens.
We speculate that the human immune system may also be more attentive 
towards beta sheets for this reason.

@@

We thank the Center for Information Technology of the University 
of Groningen for its support and for providing access to the Peregrine 
high performance computing cluster.

@@

To retrieve the data from the NCBI databases the
\verb;rentrez; R package (\cite{rentrez}) was used
that calls the NCBI website's API. To provide for a 
stable user experience for all users, 
this API limits the user to 3 calls per second.
Additionally, the API splits the result of a bigger
query into multiple pages, each of which needs one API call.
We wrote the \verb;sprentrez; package (\cite{sprentrez}) to provide for 
bigger queries of multiple (and delayed) API calls.
